Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2023-12-18 paper.tex
added intro

\definecolor{light-gray}{gray}{0.95}
\newcommand{\code}[1]{\colorbox{light-gray}{\texttt{#1}}}

@@

\address[Smith]{Schmidt Institute of Physics of the Earth of Russian Academy of
  Sciences, Moscow, 107031, Russia}
\address[MIT]{Moscow Institute of Physics and Technology,
  Dolgoprudny, 141701, Russia}
%\affiliation{\textsuperscript{3}Oil and Gas Research Institute Russian Academy
%  of Sciences (OGRI RAS) 3, Gubkina st., Moscow, 119333, Russian Federation}

\title{Another paper with lots of text and fancy pictures}

@@

\section{Theoretical background}
Let $A_1, A_2, \dots, A_n$ be pairwise disjoint subsets of a set
$A \subset \mathbb{R}^n$. We introduce a family of indicator functions
$I_n(x) : A \rightarrow \{0,1\}$ defined as:
\begin{equation}
  I_n(x) = \left\{
  \begin{array}{ll}
    1 & \quad x \in A_n \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}

@@

Another function closely related to the three-point correlation function is the
cluster function which can be defined as
\begin{equation}
  C_3^n(x_1, x_2) = \sum_{k=1}^K \langle \tilde{I}_k(x) \tilde{I}_k(x + x_1)
  \tilde{I}_k(x + x_2) \rangle
\end{equation}
where $\tilde{I}_k$ is an indicator function for $k$-th cluster and $K$ is a
total number of clusters. A cluster is a subset of $A_n$ so that any two points
in this subset can be connected with a curve belonging to that subset.

Introduce a function $M_n(x) = |\nabla I_n(x)|$ which is a so-called interface
indicator function. With this function we can define three additional
correlation functions, which are the surface-surface-surface function
($F_{sss}$), surface-surface-void function ($F_{ssv}$) and surface-void-void
function ($F_{svv}$). They are defined as
\begin{align}
  F_{sss}^n(x_1, x_2) &= \langle M_n(x) M_n(x + x_1) M_n(x + x_2) \rangle \\
  F_{ssv}^n(x_1, x_2) &= \langle M_n(x) M_v(x + x_1) I_v(x + x_2) \rangle \\
  F_{svv}^n(x_1, x_2) &= \langle M_n(x) I_v(x + x_1) I_v(x + x_2) \rangle
\end{align}
Here $I_v$ is an indicator function for the subset of the void phase
$A_v$. Unlike $S_3$ and $C_3$ functions, these functions do not have the meaning
of probability and have the units of measurement inversely proportional to
volume, surface and length, respectively (e.g. $\mu m^{-3}$, $\mu m^{-2}$ and
$\mu m^{-1}$). These functions describe the interface of the subset $A_n$ and
its spatial configuration with respect to $A_v$.

\section{Computational algorithms}
Three-point correlation function $S_3$ can be computed either in spatial domain
or in frequency domain via the convolution theorem. Suppose that a finite
countable set $A$ is defined as a multidimensional array. A point $x$ belongs to
a subset $A_n$ if and only if $A[x] = n$. The simplest algorithm below
calculates $S_3$ pointwise in spatial domain:
\begin{algorithmic}[1]
  \Procedure{$S_3$}{$A, n, x_1, x_2$}
  \State $A_n \gets I_n (A)$
  \Comment Apply $I_n$ to the input array elementwise.
  \State $A'_n \gets \mathfrak{S}_{x_1}(A_n)$
  \State $A''_n \gets \mathfrak{S}_{x_2}(A_n)$
  \State $T \gets A_n \cdot A'_n \cdot A''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}
In this algorithm the indicator function $I_n$ for the phase of interest $n$ is
applied to the input array $A$. Then the shift operator $\mathfrak{S}$ is used
to shift $A_n$ by vectors $x_1$ and $x_2$. Finally, the three arrays are
multiplied pointwise and a normalized sum of all elements of the product is
returned.

There are two common boundary conditions when applying $\mathfrak{S}$. The
first condition is to extend $A_n$ periodically (\cref{fig:s3-periodic}) when
accessing out-of-bounds array elements. In this case $\mathfrak{S}_z$ is a
circular shift operator:
\begin{equation}
  A[x] \leftarrow A[x+z \mod d]
\end{equation}
In this case $Norm(A, x_1, x_2)$ does not depend on $x_1$ and $x_2$ and
is equal to the total number of elements in $A$. The second condition is to
replace out-of-bound elements with zeros (so-called zero padding)
(\cref{fig:s3-zeros}). A formula for $Norm(A, x_1, x_2)$ in this case is in
\ref{sec:number-of-trials} and $\mathfrak{S}_z$ is defined as follows:
\begin{equation}
  A[x] \leftarrow \left\{
  \begin{array}{ll}
    A[x+z] & \quad \text{if $x+z$ is a correct index in $A$} \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}
\begin{figure*}[tp]
  \centering
  \subfigure[Periodic boundary conditions]{
    \includegraphics[width=0.45\linewidth]{images/periodic.png}
    \label{fig:s3-periodic}}
  \hfill
  \subfigure[Zero padding]{
    \includegraphics[width=0.45\linewidth]{images/zeros.png}
    \label{fig:s3-zeros}}
  \caption[]{Computation of unnormalized $S_3$ function at point $(1, 2)$ for
    one-dimensional sequence of length 8.}
  \label{fig:s3-computation}
\end{figure*}

The second approach is to compute the whole correlation map in Fourier
domain. Suppose we have a function
$f: \mathbb{R} \rightarrow \mathbb{R}$. Triple correlation for this function is
a function
\begin{equation}
  g(t_1, t_2) = \int f(\tau) f(\tau + t_1) f(\tau + t_2) d \tau
\end{equation}
We can use a well-known formula to compute Fourier image of $g$:
\begin{equation}
  \hat{g}(z_1, z_2) = \hat{f}(z_1) \hat{f}(z_2) \overline{\hat{f}(z_1 + z_2)}
\end{equation}
This algorithm has a better computational complexity compared with the previous
algorithms if the whole correlation map is needed ($O(n^2 \log n)$ vs
$O(n^3)$ where $n$ is a number of elements in the input array). The whole
correlation map, however, requires a lot of memory and is rarely
needed. Imagine, that the input array $A$ is of dimensions $500 \times 500$.
Then, considering that single precision floating point numbers are used,
$4 \cdot 2 \cdot 500^4 / 1000^3 = 500$ GB are needed to store Fourier image of
the map. Moreover, Fourier image of the whole map is needed even if only one
element of the map in spatial domain is required. The aforementioned limitations
render this approach very impractical.

@@

Algorithms for three-point surface functions are the same as for $S_3$,
containing only small modifications. The main difference is that we need to
replace an indicator function $I_n$ with an edge detection operator $E_n$
(see \ref{sec:filter}). Then the algorithms are straightforward:
\begin{algorithmic}[1]
  \Procedure{$F_{sss}$}{$A, n, x_1, x_2$}
  \State $A_n \gets E_n (A)$
  \State $A'_n \gets \mathfrak{S}_{x_1}(A_n)$
  \State $A''_n \gets \mathfrak{S}_{x_2}(A_n)$
  \State $T \gets A_n \cdot A'_n \cdot A''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure

  \Procedure{$F_{ssv}$}{$A, n, x_1, x_2$}
  \State $A_n \gets E_n (A)$
  \State $A_{(void)} \gets I_{(void)} (A)$
  \State $A'_n \gets \mathfrak{S}_{x_1}(A_n)$
  \State $A'_{(void)} \gets \mathfrak{S}_{x_2}(A_{(void)})$
  \State $T \gets A_n \cdot A'_n \cdot A'_{(void)}$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure

  \Procedure{$F_{svv}$}{$A, n, x_1, x_2$}
  \State $A_n \gets E_n (A)$
  \State $A_{(void)} \gets I_{(void)} (A)$
  \State $A'_{(void)} \gets \mathfrak{S}_{x_1}(A_{(void)})$
  \State $A''_{(void)} \gets \mathfrak{S}_{x_2}(A_{(void)})$
  \State $T \gets A_n \cdot A'_{(void)} \cdot A''_{(void)}$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}

Input images must have appropriately high resolution for edge detection filter
to work correctly. A criterion for ``goodness'' of an image can be expressed
with $C_{\alpha}$ parameter \cite{samarin2023robust}. Let $\alpha \in [0, 1]$
and $f(x)$ be (band-limited) input image ($x \in \mathbb{R}^2$ or
$\mathbb{R}^3$). We define $C_\alpha$ as follows:
\begin{equation}
  \begin{aligned}
    f_0(x) &= f(x) - \langle f(x) \rangle \\
    C_\alpha &= \frac{\int_{-\alpha\omega}^{\alpha\omega} |\hat{f_0}(z)|^2
      dz}{\int_{-\omega}^{\omega} |\hat{f_0}(z)|^2 dz}
  \end{aligned}
\end{equation}
where $\hat{f_0}$ is a Fourier transform of $f_0$ and $\omega$ is a folding
frequency which depends only on resolution of the image. If a criterion
$C_{0.5} > 0.97$ holds then the image is suitable for application of
aforementioned algorithms for calculation of surface functions.

To compute $C_3$ function we need to use a connected-component labeling
algorithm \cite{4728561,PhysRevB.14.3438} to label clusters. Then, again, a
slightly modified algorithm for $S_3$ is used.
\begin{algorithmic}[1]
  \Procedure{$C_3$}{$A, n, x_1, x_2$}
  \State $A_n \gets I_n (A)$
  \State $C_n \gets \mathfrak{C}(A_n)$
  \State $C'_n \gets \mathfrak{S}_{x_1}(C_n)$
  \State $C''_n \gets \mathfrak{S}_{x_2}(C_n)$
  \State $T \gets C_n \odot C'_n \odot C''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}
Here $\mathfrak{C}$ is connected-component labeling operator and $\odot$ is
defined as follows:
\begin{equation}
  x \odot y = \left\{
  \begin{array}{ll}
    1 & \quad x = y \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}

\section{Verification}
To verify correctness of computation of $S_3$ and $C_3$ functions we use the
following simple relations:
\begin{align}
  S_3^n (x, 0) = S_3^n (0, x) &= S_2^n(x) \\
  C_3^n (x, 0) = C_3^n (0, x) &= C_2^n(x) \\
  \lim_{\substack{x_1 \to \infty \\ x_2 \to \infty}} S_3^n (x_1, x_2) &= \phi_n^3
\end{align}
Here $\phi_n$ is a fraction of phase $n$ in $A$.

There are only few sets for which analytic expressions for $S_2$ are known. One
of those is a set consisting of overlapping balls with a fixed radius $R$ and
centers generated with Poisson process with parameter $\lambda$. In this case
the expression for $S_2$ is following:
\begin{equation}
  \begin{aligned}
    S_2(r, R) &= \exp(-\frac{4}{3}\pi\lambda R^3 f(r, R)) \\
    f(r, R) &= \left\{
    \begin{array}{ll}
      1 + \frac{3}{4} \frac{r}{R} - \frac{1}{16} (\frac{r}{R})^3 & \quad r < 2R \\
      2 & \quad \text{otherwise}
    \end{array}
    \right.
  \end{aligned}
  \label{eq:s2-balls}
\end{equation}

On \cref{fig:balls} there is an intersection of such a set with $R = 0.02$ and
$\lambda=5000$ and a cube $[0, 1]^3$. Because $R \ll 1$, we can use
\cref{eq:s2-balls} to verify our computations what we do on
\cref{fig:balls-s3-comparison}.
\begin{figure*}[tp]
  \centering
  \subfigure[Realization of overlapping balls]{
    \includegraphics[width=0.45\linewidth]{images/balls.png}
    \label{fig:balls}}
  \hfill
  \subfigure[A plot of $S_3(0, r)$ computed with our algorithm and theoretical
    formula.]{
    \includegraphics[width=0.45\linewidth]{images/balls-s3.png}
    \label{fig:balls-s3-comparison}}
  \caption[]{A comparison of calculated and theoretical values of $S_3$ function
    for overlapping balls.}
  \label{fig:s3-verification}
\end{figure*}

To test $F_{sss}$ function we use a recently developed approach which computes
that function for sets defined by inequation $f(x) \le T$, $x \in \mathbb{R}^3$
with help of automatic differentiation \cite{postnicov20232}. We choose $f$ as
\begin{equation}
  f(x, y, z) = y^2 + z^2 - a x^2 (b^2 - x^2)
\end{equation}
and parameters $a = 10$, $b = 0.45$, $T = 10^{-2}$, so inequation
$f(x, y, z) \le T$ gives us a dumbbell-like object (\cref{fig:sss-dumbbell}).

>>>>

\definecolor{light-gray}{gray}{0.95}
\newcommand{\code}[1]{\colorbox{light-gray}{\texttt{#1}}}
\newcommand{\highlight}[1]{{\color{red}{#1}}} % convinient for revised version

@@

\address[Smith]{Schmidt Institute of Physics of the Earth of Russian Academy of
  Sciences, Moscow, 107031, Russia}
\address[MIT]{Moscow Institute of Physics and Technology,
  Dolgoprudny, 141701, Russia}

\title{Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects}

@@

\section{Introduction}
Structures, or spatial arrangements of matter and energy, including some fields (e.g., velocity or pressure) are ubiquitous in research applications and frequently require description for subsequent analysis. Examples of such objects for studies include galaxy formations \cite{springel2006}, immiscible multiphase fluid flow patterns \cite{hopkins2015new,balashov2021}, rock and soil samples \cite{rozenbaum2014,karsanina2015,ledesma2018,chen2020super,prokhorov2021digital,vogel2010}, food specimens \cite{derossi2019,nagdalian2021}, biological tissue \cite{park2020}, vortexes during flow \cite{gorbunova2016precessing} and interstellar turbulence \cite{portillo2018developing}. The scale of interest may span orders of magnitude from angstroms and nm for materials \cite{garum2020,gerke2021,khlyupin2023molecular} to millions of light-years for star clusters \cite{takada2003three,hopkins2013stars}. The studied structure can be non-static -- it will  change with time, and this dynamics is usually also of great interest \cite{jiao2013,fomin2023soil}. Thus, description of the structure is of utmost importance for numerous applications:
\begin{enumerate}
  \item Elucidation of structure--property relationships in the form of theoretical expression \cite{Torquato_book,Sahimi_book}, e.g., rigorous bounds and such;
  \item Prediction of material physical properties using machine (deep) learning \cite{obayashi2018persistence,kamrava2020linking,roding2020predicting};
  \item Description of complex structures and fields for taxonomy, comparison and analysis \cite{takada2003three,hopkins2013stars,shivashankar2015felix,saadatfar2017pore,portillo2018developing,KarsaninaEJSS,PNM_Morse,khlyupin2023molecular}, which may include temporal dynamics or surface evolution \cite{jiao2013,PhysRevE.92.023301,prokhorov2022,chen2022,fomin2023soil};
  \item Morphological representativity \cite{capek2009,rozenbaum2014,gerke2019tensor} and spatial stationarity analysis \cite{REVpaper,LavrukhinPRE};
  \item Stochastic reconstructions from limited amount of input data \cite{Adler_recon,Y-T,tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,rozanski2023} in the form of limited dimension structures or correlation functions;
  \item Compress structural information for storage, analysis, training (feature extraction), and retrieval (by means of stochastic reconstructions) \cite{jiao2007,SciRep1,Havelka,KarsaninaEJSS};
  \item Multiscale image fusion that is usually necessary due to trade-off between field-of-view and imaging resolution \cite{SciRep1,Geoderma2018,chen2020super,karimpouli2022}.
\end{enumerate}

The number of possible descriptors to quantify spatial structure developed within different disciplines is stunning, for example: phase ratio or porosity \cite{anovitz2015characterization}, surface area, radial distribution functions \cite{zimm1948scattering,becker2010radial}, Minkowski functionals \cite{vogel2010} and tensors \cite{schroder2011minkowski}, correlation functions \cite{Torquato_book}, persistence diagrams \cite{shivashankar2015felix,saadatfar2017pore,obayashi2018persistence}. The evolution of such metrics was mainly induced by the ease of measurement and interpretation -- this is startling to observe holy wars on the question which metric is superior to flare up occasionally, e.g.,\cite{vogel2022holistic,yudina2023dual}. In fact, we need a single universal descriptor defined by only one characteristic -- information content. The latter is an ability of the metric to describe the structure at hand with 100$\%$ meaning the ability to fully recover the structure back from this metric. Thus, universal descriptor has to possess following key properties: (1) high information content, (2) easy to assess, (3) high compression (i.e., descriptor is much "smaller" than the structure itself).

The large portion of the research papers mentioned above \cite{rozenbaum2014,karsanina2015,ledesma2018,derossi2019,portillo2018developing,takada2003three,hopkins2013stars,jiao2013,Torquato_book,roding2020predicting,KarsaninaEJSS,capek2009,gerke2019tensor,REVpaper,LavrukhinPRE,Adler_recon,Y-T,tahmasebiPRL,Euras2012,EPL2,karsaninaPRL,jiao2007,SciRep1,Havelka,Geoderma2018} is based on so-called correlation functions (CFs). These functions define a probability that some events are occurring withing the structure (with some special cases of surface functions), for example, 2-point probability $S_2$ measures the probability that both ends of the line segment are within the same phase, 2-point cluster function $C_2$ - that both ends of the line segment are within the same interconnected cluster of same phase, and lineal path $L_2$ measures the probability that the whole segment lies within the phase (see \highlight{Fig.1}). Compared to other descriptors, CFs possess a number of features making them, in our opinion, a perfect candidate for universal descriptor:
\begin{enumerate}
  \item They contain the majority of other metrics inside. For example, porosity is $S_2(0)$ (and at zero correlation length of the majority of correlation functions except for surface ones), surface area is represented by in the derivative at $S_2(0)$ \cite{debye1957scattering} or the value of $F_{ss}(0)$ (or $F_{sss}(0)$). Connectivity is described by cluster functions $C_2$ or $C_3$. Pore-size distribution is related to pore-size function and partially by $L_2$ and $S_3$. Surface-void functions are interconnected with mean curvature and Euler characteristic \cite{ma2020generation}. All together, the classic set of correlation functions describes both the geometry and topology of the structure at hand;
  \item While the extent to which such classical set describes any arbitrary structure is not yet known, compared to other complex descriptors  where are known pathways to assess the information content of correlation functions -- there is a methodology for $S_2$ \cite{Gommes2} and approaches for other functions \cite{Degeneraty.045306,cherkasov2023towards};
  \item In addition to a possibility to compute correlation functions using images or other structural data, i.e., similar to all other descriptors, some CFs can be also determined experimentally \cite{debye1957scattering,barrall1992nmr,dietrich1995scattering,li2018accurate};
  \item Correlation functions can be scaled \cite{karsaninaPRL} or manipulated, for example, to add functions into a single one \cite{moctezuma2002};
  \item As opposed to scalar metrics such as porosity, surface area or Minkowski functionals, CFs can be computed in directions and describe anisotropic structures;
  \item Numerous rigorous bounds to obtain thermal, elastic, electromagnetic and transport properties are known, among which the most known (and least useful due to the impractically huge gap between the bounds) are Hashin-Strikman bounds \cite{hashin1963variational}. Other expressions based on integration of higher order $S_n$ are known and were found to be close to direct simulations \cite{brown1955solid,beran1965use,milton1981bounds,hlushkou2015effective};
  \item One can control both the size of the CFs-based descriptor and the information content via correlation length, number of functions in the set (including directions \cite{EPL2}), and by the order of functions (e.g., 2- and 3-point versions, or even more points).
\end{enumerate}
This list highlights that correlations functions are indeed possess necessary properties so satisfy key criteria stated above.

However, while 2-point CFs are computed and used in multitude of studies, higher order functions, for example, 3-point correlation functions are not that widely used. The reason for this is straightforward -- 3-point correlation functions are more involved and, thus, require much more computational resources to assess. It has been shown that the amount of additional information content with each increase in number of points $n$ decreases \cite{yao1993high,Gommes2}. The trade-off between additional information and computational burden is still not clear. Nonetheless, the computation of 3-point statistics is known to be advantageous at least in the number of applications as related to cosmology, astrophysics and turbulence patterns analysis as they can probe non-Gaussian signatures in the distribution of matter and energy, which are not captured in 2-point counterparts \cite{TakadaJain,hopkins2013stars,gorbunova2016precessing,yoo2022non}.

In this contribution we continue our initiative to develop a simple, efficient and open-source solution to compute all classical correlation functions on CPU and GPU architectures -- \code{CorrelationFunctions.jl} \cite{CFsjlpaper}. Here we build upon previous results in computations of higher order functions with n > 2, most notably by Malmir et al. \cite{malmir2018}, and our aim is to develop robust and computationally efficient methods. In addition, we also considered the case of anisotropic structures and how to describe them with 3-point correlation functions similar to directional 2-point CFs \cite {10.1063/1.4867611,EPL1}.

\highlight{The paper is organized as follows:}

\section{Methods}
\subsection{3-point correlation functions}
Let $A_1, A_2, \dots, A_n$ be pairwise disjoint subsets of a set
$A \subset \mathbb{R}^n$, where the index can be interpreted as a phase on the image $A$. We introduce a family of indicator functions
$I_n(x) : A \rightarrow \{0,1\}$ defined as:
\begin{equation}
  I_n(x) = \left\{
  \begin{array}{ll}
    1 & \quad x \in A_n \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}

@@

Another function closely related to the three-point correlation function is the
cluster function which can be defined as:
\begin{equation}
  C_3^n(x_1, x_2) = \sum_{k=1}^K \langle \tilde{I}_k(x) \tilde{I}_k(x + x_1) \tilde{I}_k(x + x_2) \rangle
\end{equation}
where $\tilde{I}_k$ is an indicator function for $k$-th cluster and $K$ is a
total number of clusters. A cluster is a subset of $A_n$ so that any two points
in this subset can be connected with a curve belonging to that subset.

Introduce a function $M_n(x) = |\nabla I_n(x)|$ which is a so-called interface
indicator function. With this function we can define three additional
correlation functions, which are the surface-surface-surface function
($F_{sss}$), surface-surface-void function ($F_{ssv}$) and surface-void-void
function ($F_{svv}$). They are defined as:
\begin{align}
  F_{sss}^n(x_1, x_2) &= \langle M_n(x) M_n(x + x_1) M_n(x + x_2) \rangle \\
  F_{ssv}^n(x_1, x_2) &= \langle M_n(x) M_v(x + x_1) I_v(x + x_2) \rangle \\
  F_{svv}^n(x_1, x_2) &= \langle M_n(x) I_v(x + x_1) I_v(x + x_2) \rangle
\end{align}
Here $I_v$ is an indicator function for the subset of the void phase
$A_v$. Unlike $S_3$ and $C_3$ functions, these functions do not have the meaning
of probability and have the units of measurement inversely proportional to
volume, surface and length, respectively (e.g. $\mu m^{-3}$, $\mu m^{-2}$ and
$\mu m^{-1}$). These functions describe the interface of the subset $A_n$ and
its spatial configuration with respect to $A_v$.

\subsection{Computational algorithms}
Three-point correlation function $S_3$ can be computed either in spatial domain
or in frequency domain via the convolution theorem. The simplest algorithm below
calculates $S_3$ pointwise in spatial domain:
\begin{algorithmic}[1]
  \Procedure{$S_3$}{$A, n, x_1, x_2$}
  \State $A_n \gets I_n (A)$
  \State $A'_n \gets \mathfrak{S}(A_n, x_1)$
  \State $A''_n \gets \mathfrak{S}(A_n, x_2)$
  \State $T \gets A_n \cdot A'_n \cdot A''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}
In this algorithm the indicator function $I_n$ for the phase of interest $n$ is
applied to the input array $A$. Then the shift operator $\mathfrak{S}$ is used
to shift $A_n$ by vectors $x_1$ and $x_2$. Finally, the three arrays are
multiplied pointwise and normalized sum of all elements of the product is
returned.

There are two common boundary conditions when applying $\mathfrak{S}$. The first
condition is to extend $A_n$ periodically (\cref{fig:s3-periodic}) when
accessing out-of-bounds array elements. This is the same as circular-shifting
$A_n$. In this case $Norm(A, x_1, x_2)$ does not depend on $x_1$ and $x_2$ and
is equal to the total number of elements in $A$. The second condition is to
replace out-of-bound elements with zeros (so-called zero padding)
(\cref{fig:s3-zeros}). A formula for $Norm(A, x_1, x_2)$ in this case is in
\ref{sec:number-of-trials}.
\begin{figure*}[tp]
  \centering
  \subfigure[Periodic boundary conditions]{
    \includegraphics[width=0.45\linewidth]{images/periodic.png}
    \label{fig:s3-periodic}}
  \hfill
  \subfigure[Zero padding]{
    \includegraphics[width=0.45\linewidth]{images/zeros.png}
    \label{fig:s3-zeros}}
  \caption[]{Computation of unnormalized $S_3$ function at point $(1, 2)$ for
    one-dimensional sequence of length 8.}
  \label{fig:s3-computation}
\end{figure*}

The second approach is to compute the whole correlation map in Fourier
domain. Suppose we have a function
$f: \mathbb{R} \rightarrow \mathbb{R}$. Triple correlation for this function is:
\begin{equation}
  g(t_1, t_2) = \int f(\tau) f(\tau + t_1) f(\tau + t_2) d \tau.
\end{equation}
We can use a well-known formula to compute Fourier image of $g$:
\begin{equation}
  \hat{g}(z_1, z_2) = \hat{f}(z_1) \hat{f}(z_2) \overline{\hat{f}(z_1 + z_2)}
\end{equation}
This algorithm has a better computational complexity compared with the previous
algorithms if the whole correlation map is needed ($O(n^2 \log n)$ vs
$O(n^3)$ where $n$ is a number of elements in the input array). The whole
correlation map, however, requires a lot of memory and is rarely
needed. Imagine, that the input array $A$ is of dimensions $500 \times 500$.
Then, considering that single precision floating point numbers are used,
$4 \cdot 2 \cdot 500^4 / 1000^3 = 500$ GB are needed to store Fourier image of
the map. Moreover, Fourier image of the whole map is needed even if only one
element of the map in spatial domain is required. The aforementioned limitations
render this approach impractical.

@@

Algorithms for three-point surface functions are the same as for $S_3$,
containing only small modifications. The main difference is that we need to
replace an indicator function $I_n$ with an edge detection operator $E_n$
(see \ref{sec:filter}). Then the algorithms are straightforward:
\begin{algorithmic}[1]
  \Procedure{$F_{sss}$}{$A, n, x_1, x_2$}
  \State $A_n \gets E_n (A)$
  \State $A'_n \gets \mathfrak{S}(A_n, x_1)$
  \State $A''_n \gets \mathfrak{S}(A_n, x_2)$
  \State $T \gets A_n \cdot A'_n \cdot A''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure

  \Procedure{$F_{ssv}$}{$A, n, x_1, x_2$}
  \State $A_n \gets E_n (A)$
  \State $A_{(void)} \gets I_{(void)} (A)$
  \State $A'_n \gets \mathfrak{S}(A_n, x_1)$
  \State $A'_{(void)} \gets \mathfrak{S}(A_{(void)}, x_2)$
  \State $T \gets A_n \cdot A'_n \cdot A'_{(void)}$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure

  \Procedure{$F_{svv}$}{$A, n, x_1, x_2$}
  \State $A_n \gets E_n (A)$
  \State $A_{(void)} \gets I_{(void)} (A)$
  \State $A'_{(void)} \gets \mathfrak{S}(A_{(void)}, x_1)$
  \State $A''_{(void)} \gets \mathfrak{S}(A_{(void)}, x_2)$
  \State $T \gets A_n \cdot A'_{(void)} \cdot A''_{(void)}$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}

Input images must have appropriately high resolution for edge detection filter
to work correctly. A criterion for ``goodness'' of an image can be expressed
with $C_{\alpha}$ parameter \cite{samarin2023robust}. Let $\alpha \in [0, 1]$
and $f(x)$ be (band-limited) input image ($x \in \mathbb{R}^2$ or
$\mathbb{R}^3$). We define $C_\alpha$ as follows:
\begin{equation}
  \begin{aligned}
    f_0(x) &= f(x) - \langle f(x) \rangle \\
    C_\alpha &= \frac{\int_{-\alpha\omega}^{\alpha\omega} |\hat{f_0}(z)|^2
      dz}{\int_{-\omega}^{\omega} |\hat{f_0}(z)|^2 dz}
  \end{aligned}
\end{equation}
where $\hat{f_0}$ is a Fourier transform of $f_0$ and $\omega$ is a folding
frequency which depends only on resolution of the image. If a criterion
$C_{0.5} > 0.97$ holds then the image is suitable for application of
aforementioned algorithms for calculation of surface functions, otherwise image's resolution
can be improved with the help of interpolation prior to computations \cite{samarin2023robust}.

To compute $C_3$ function we need to use a connected-component labeling
algorithm \cite{4728561,PhysRevB.14.3438} to label clusters. Then, again, a
slightly modified algorithm for $S_3$ is used to assess cluster function:
\begin{algorithmic}[1]
  \Procedure{$C_3$}{$A, n, x_1, x_2$}
  \State $A_n \gets I_n (A)$
  \State $C_n \gets \mathfrak{C}(A_n)$
  \State $C'_n \gets \mathfrak{S}(C_n, x_1)$
  \State $C''_n \gets \mathfrak{S}(C_n, x_2)$
  \State $T \gets C_n \odot C'_n \odot C''_n$
  \State \textbf{return} $Sum(T) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}
Here $\mathfrak{C}$ is connected-component labeling operator and $\odot$ is
defined as follows:
\begin{equation}
  x \odot y = \left\{
  \begin{array}{ll}
    1 & \quad x = y \\
    0 & \quad \text{otherwise}.
  \end{array}
  \right.
\end{equation}

\section{Results}
\subsection{Verification}
To verify correctness of computation of $S_3$ and $C_3$ functions we use the
following simple relations:
\begin{align}
  S_3^n (x, 0) = S_3^n (0, x) &= S_2^n(x) \\
  C_3^n (x, 0) = C_3^n (0, x) &= C_2^n(x) \\
  \lim_{\substack{x_1 \to \infty \\ x_2 \to \infty}} S_3^n (x_1, x_2) &= \phi_n^3
\end{align}
Here $\phi_n$ is a fraction of phase $n$ in $A$.

There are only few sets for which analytic expressions for $S_2$ are known. One
of those is a set consisting of overlapping balls with a fixed radius $R$ and
centers generated with Poisson process with parameter $\lambda$. In this case
the expression for $S_2$ is following:
\begin{equation}
  \begin{aligned}
    S_2(r, R) &= \exp(-\frac{4}{3}\pi\lambda R^3 f(r, R)) \\
    f(r, R) &= \left\{
    \begin{array}{ll}
      1 + \frac{3}{4} \frac{r}{R} - \frac{1}{16} (\frac{r}{R})^3 & \quad r < 2R \\
      2 & \quad \text{otherwise}.
    \end{array}
    \right.
  \end{aligned}
  \label{eq:s2-balls}
\end{equation}

On \cref{fig:balls} there is an intersection of such a set with $R = 0.02$ and
$\lambda=5000$ and a cube $[0, 1]^3$. Because $R \ll 1$, we can use
\cref{eq:s2-balls} to verify our computations -- this is exactly what we present on
\cref{fig:balls-s3-comparison}.
\begin{figure*}[tp]
  \centering
  \subfigure[Realization of overlapping balls]{
    \includegraphics[width=0.45\linewidth]{images/balls.png}
    \label{fig:balls}}
  \hfill
  \subfigure[A plot of $S_3(0, r)$ computed with our algorithm and theoretical
    formula.]{
    \includegraphics[width=0.45\linewidth]{images/balls-s3.png}
    \label{fig:balls-s3-comparison}}
  \caption[]{A comparison of calculated and theoretical values of $S_3$ function
    for overlapping balls.}
  \label{fig:s3-verification}
\end{figure*}

To test $F_{sss}$ function we use a recently developed approach which computes
that function for sets defined by inequality $f(x) \le T$, $x \in \mathbb{R}^3$
with help of automatic differentiation \cite{postnicov20232}. We choose $f$ as:
\begin{equation}
  f(x, y, z) = y^2 + z^2 - a x^2 (b^2 - x^2)
\end{equation}
and parameters $a = 10$, $b = 0.45$, $T = 10^{-2}$, so inequation
$f(x, y, z) \le T$ gives us a dumbbell-like object (\cref{fig:sss-dumbbell}).

The test is performed by firstly evaluating the inequation in a cube
$[-1, 1]^3$ and obtaining precise values of $F_{sss}$ and secondly computing
$F_{sss}$ with our algorithm, giving a discretized $400 \times 400 \times 400$
pixels wide image as the input. $F_{sss}$ function is evaluated in $XY$
plane. Pointwise relative error of $F_{sss}$ is on
\cref{fig:sss-dumbbell-error}. Maximal relative error in the area distanced
farther than 0.1 (20 pixels) from curves of discontinuity (see
\cref{fig:sss-dumbbell-precise}) is no more than 20\%.
\begin{figure*}[tp]
  \centering
  \subfigure[The dumbbell-like object used in testing of $F_{sss}$ computation]{
    \includegraphics[width=0.45\linewidth]{images/dumbbell.png}
    \label{fig:sss-dumbbell}}
  \hfill
  \subfigure[Relative error of computation]{
    \includegraphics[width=0.45\linewidth]{images/dumbbell-sss-err.png}
    \label{fig:sss-dumbbell-error}}
  \vskip\baselineskip
  \subfigure[$F_{sss}$: precise. The correlation function is undefined (has type
    II discontinuity) at curves A, B and C.]{
    \includegraphics[width=0.45\linewidth]{images/dumbbell-sss-precise.png}
    \label{fig:sss-dumbbell-precise}}
  \hfill
  \subfigure[$F_{sss}$: calculated with our approach]{
    \includegraphics[width=0.45\linewidth]{images/dumbbell-sss-julia.png}}
  \caption[]{A comparison of $F_{sss}$ function calculated using our approach
    with its precise values.}
  \label{fig:sss-verification}
\end{figure*}
